D.I.A.G.R.A.M.: Development of Image Analysis for Graph Recognition And Modeling

### 2025-06-29 doc/article/article.tex

\subsection{Classifier}
\label{exp:classifier}
The first architecture explored for the Classifier module was a Convolutional Neural Network (CNN) following AlexNet's architecture, but smaller in size.
It had two Convolutional layers, each followed by a batch normalization and then by a max pooling layers. The first layer had only 4 filters, while the second one had 8.
The output of the second layer was then flattened and passed through a fully connected layer which downsized the features to 128, followed by a ReLU activation function. Other two fully connected layer brought the number of output features down to the number of classes.
Then, a softmax activation function was applied to the output layer, in order to obtain the probabilities of each class.
The model was trained on our own dataset, with only three possible output classes: 'flowchart', 'graph' and 'other'; it was trained for 10 epochs, with an AdamW optimizer (with a learning rate of 1e-3) and a standard CrossEntropy Loss function.
\\

@@

Judging this to be a problem in architecture size, we then tried adding another CNN layer with 16 filters; this one too followed by a batch normalization and a max pooling layer.
The rest of the net was untouched; we also decreased the learning rate to 1e-4, in order to avoid overshooting the minimum of the loss function.
The results, however, didn't show any kind of improvement, with the loss function still being around 0.9.
\\

@@

The original model was trained over the COCO dataset \cite{coco}, which comprised 91 classes. It should be highlighted that the original training dataset was not in any way related to diagrams: with an efficient finetune, the network was able to swiftly change to our task. \\

The model employed was tested in two different ways: as zero-shot, after a 10-epoch training and after a 100-epoch training. The best-behaving one was, of course, the latter (to see the results of the different stages, see \textbf{Figures}~\ref{fig:bbox1} -~\ref{fig:bbox2} -~\ref{fig:bbox3}). In Table~\ref{tab:map} and Table~\ref{tab:mar} we show the Mean Average Precision and Recall of the model used.

\begin{figure}
	\centering
	\includegraphics[width=200pt, height=150pt]{bbox1.jpg}
	\caption{The model employed in a zero-shot fashion.}
	\label{fig:bbox1}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=200pt, height=150pt]{bbox2.jpg}
	\caption{The model employed after 10 epochs of training.}
	\label{fig:bbox2}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=180pt, height=150pt]{bbox3.png}
	\caption{The model employed after 100 epochs of training. }
	\label{fig:bbox3}
\end{figure}

@@

The solution, as stated before, was to finetune the already working RCNN in order for it to also extract the bounding boxes of both tails and heads of the arrow.

@@

>>>>

\subsection{Classifier}
\label{exp:classifier}
The first architecture explored for the Classifier module was a Convolutional Neural Network (CNN) following AlexNet's architecture, but smaller in size.
It had two Convolutional layers, each followed by a batch normalization and then by a max pooling layers. The first layer had only 4 filters, while the second one had 8.
The output of the second layer was then flattened and passed through a fully connected layer which downsized the features to 128, followed by a ReLU activation function. Other two fully connected layer brought the number of output features down to the number of classes.
Then, a softmax activation function was applied to the output layer, in order to obtain the probabilities of each class.
The model was trained on our own dataset, with only three possible output classes: 'flowchart', 'graph' and 'other'; it was trained for 10 epochs, with an AdamW optimizer (with a learning rate of $0.001$) and a standard CrossEntropy Loss function.
\\

@@

Judging this to be a problem in architecture size, we then tried adding another CNN layer with 16 filters; this one too followed by a batch normalization and a max pooling layer.
The rest of the net was untouched; we also decreased the learning rate to $0.0001$, in order to avoid overshooting the minimum of the loss function.
The results, however, didn't show any kind of improvement, with the loss function still being around 0.9.
\\

@@

The original model was trained over the COCO dataset \cite{coco}, which comprised 91 classes. It should be highlighted that the original training dataset was not in any way related to diagrams: with an efficient finetune, the network was able to swiftly change to our task. The finetuning is done with a learning rate of $0.005$ \\

The model employed was tested in three different ways: as zero-shot, after a 10-epoch training and after a 100-epoch training. The best-behaving one was, of course, the latter (to see the results of the different stages, see \textbf{Figures}~\ref{fig:bbox1} -~\ref{fig:bbox2} -~\ref{fig:bbox3} -~\ref{fig:fc}). In Table~\ref{tab:map} and Table~\ref{tab:mar} we show the Mean Average Precision and Recall of the model used.

\begin{figure}[H]
	\centering
	\includegraphics[width=200pt, height=150pt]{bbox1.jpg}
	\caption{The model employed in a zero-shot fashion.}
	\label{fig:bbox1}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=200pt, height=150pt]{bbox2.jpg}
	\caption{The model employed after 10 epochs of training.}
	\label{fig:bbox2}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=180pt, height=150pt]{bbox3.png}
	\caption{The model employed after 100 epochs of training. More bounding boxes can be seen because we also added arrow's tails and heads.}
	\label{fig:bbox3}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=180pt, height=150pt]{fc.png}
	\caption{The bounding boxes of a flowchart diagram, on the 100-epoch model.}
	\label{fig:fc}
\end{figure}

@@

The solution, as stated before, was to finetune the already working Faster R-CNN in order for it to also extract the bounding boxes of both tails and heads of the arrow.

@@

TODO: Nodo extra
